---
title: "EDS 223 Homework Assignment 1"
subtitle: "Map Making Practise"
author: "Kelly Loo"
date: 2026-10-06
execute: 
  eval: true
format:
  pdf:
    toc: false
    mainfont: Times New Roman
options: 
  chunk_output_type: console
---

## Load libraries and data

We start off by loading in the necessary libraries.
```{r}
#| output: False
library(tidyverse)
library(stars)
library(here)
library(sf)
library(tmap)
```

Then, load in our data with `st_read()` as it is a vector.
```{r}
#| output: False
ejscreen <- st_read(here("data", "ejscreen",
"EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))
```

## Filtering

For this assignment, I will be analysing San Diego County in California. Therefore, I'm taking a subsect of data for the county.

```{r}
sandiego <- ejscreen |> 
    filter(STATE_NAME == "California") |> 
    filter(CNTY_NAME == "San Diego County")
```

<!-- {{< pagebreak >}} -->

## Investigating

Is there a correlation between cancer risks from carcinogens in ambient outdoor air and proximity to traffic in San Diego County?

<!-- Plain map of only air toxics cancer risk -->
```{r}
#| warning: false
tm_shape(sandiego , bbox = c(-117.7, 32.3, -116, 33.6)) +
    tm_polygons(
        "CANCER",
        fill.legend = tm_legend(title = "Cancer risk (^percentile)", 
        position = tm_pos_out("right")),
        col_alpha = 0
    ) +
    tm_title(text = "San Diego County: Air toxics cancer risk") +
    tm_compass(position = c("bottom","left")) +
    tm_graticules() +
    tm_scalebar(position = c("bottom","left"))
```
_^According to EJScreen, the air toxics cancer risk indicator was calculated to a statistical percentile. It results in a range from 0 to 100%, with the 35th percentile corresponding to a raw value of 30._

A lifetime of inhaling airborne carcinogens leads to an increased cancer risk. This variable in the EJScreen dataset displays these estimations broken down by census tract. Based on the map, there is a concentration of areas with higher air toxics cancer risk near the border in areas such as Chula Vista.

Engine exhaust is an example of a carcinogen. I explore this variable in the next map.

```{r}
#| warning: false
tm_shape(sandiego , bbox = c(-117.7, 32.3, -116, 33.6)) +
    tm_polygons(
        "CANCER",
        fill.legend = tm_legend(title = "Cancer risk (^percentile)", 
        position = tm_pos_out("right")),
        col_alpha = 0
    ) +
    tm_symbols(
        size = "PTRAF",
        fill = "green",
        size.legend = tm_legend(title = "Proximity to traffic (m)", 
        position = tm_pos_out("right")),
        fill_alpha = 0.25,
        col_alpha = 0.1
    ) +
    tm_title(text = "San Diego County: Air Toxics Cancer Risk and \n
    Proximity to Traffic") +
    tm_compass(position = c("bottom","left")) +
    tm_graticules() +
    tm_scalebar(position = c("bottom","left"))
```
_^See note above_

San Diego has three major highways: Interstate 5, 15, and 8 which runs along the coast, inland, and parallel to the border respectively. These can be seen with the density of green. It is no surprise that areas that have proximity to traffic also tends to have higher air toxic cancer risk. Majority of the population in San Diego live near the coast or downtown, where commute via car is common. According to the map, areas with proximity to traffic and higher risk of cancer from air toxics are near the coast and downtown. This could be a call for an increase in public transit and/or to encourage electric vehicle usage.
